### 2021-12-17 text/jsm/jsm_paper.tex
add proof + more citations

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021} and improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction. We extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

@@

Next, we studied the thresholding method from a theoretical perspective, recovering in precise mathematical terms phenomena revealed in the empirical analysis (plus several others). We worked in a simplified Gaussian setting. Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
\begin{equation}\label{theoretical_model}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i; \quad
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
p_i \sim \textrm{Bern}(\pi); \quad
\ep_i, \tau_i \sim N(0,1),
\end{equation}
where $p_i, \tau_i,$ and $\ep_i$ are independent. For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e. $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We begin by deriving the limit of $\hat{\beta}^m_1$: 
\begin{proposition}\label{prop:convergence}
	The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
	\begin{equation}\label{thresh_est_intercepts}
	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
	\end{equation} where 
	$$
	\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
\end{proposition}
The function $\gamma: \R^4 \to \R$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \R^4 \to \R$ of $\hat{\beta}^m_1$ is given by
$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
Having derived an exact expression for  the asymptotic relative bias of  $\hat{\beta}^m_1$, we can prove several results about this quantity. For the sake of simplicity, we consider a slightly specialized setting in which we set $\pi$ to to a specific value (see Appendix \ref{sec:appendix_theory} for details; proofs and detailed statements likewise deferred to Appendix \ref{sec:appendix_theory}).

@@

To answer this question, we introduce GLM-EIV (GLM-based errors-in-variables), a method that models the entire data-generating process.

\subsection{Model}\label{sec:model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_2}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as $
p_i \sim \textrm{Bern}(\pi),$ where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}), (\ref{glmeiv_model_2}), and the marginal distribution of $p_i$ define the negative binomial GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors. See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts.

@@

We see from (\ref{marginal_density}) that the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities. Additionally, the GLM-EIV model is a generalization the simple errors-in-variables model (when the predictor is binary); the latter is defined as follows:
\begin{equation}\label{classical_eiv}
y_i = \beta_0 + \beta_1 x^*_i + \ep_i; \quad
x_i = x^*_i + \tau_i,
\end{equation}
where, $x^*_i \sim \textrm{Bern}(\pi), \ep_i, \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent. GLM-EIV extends (\ref{classical_eiv}) in at least three directions: first, GLM-EIV allows $y_i$ and $x_i$ to follow (possibly different) exponential family distributions; second, GLM-EIV allows $y_i$ and $x_i$ to be related to $x^*_i$ through (possibly different) nonlinear link functions; and finally, GLM-EIV allows confounders $z_i$ to impact both $x_i$ and $y_i$. In such settings $x_i$ and $y_i$ are dependent \textit{even after conditioning on $x^*_i$}, enabling GLM-EIV to capture more complex dependence relationships between $x_i$ and $y_i$ than is possible in (\ref{classical_eiv}) or other standard measurement error models.

@@

Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV and the thresholding method to analyze the entire Gasperini and Xie datasets. (We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good threshold when the gRNA count distribution is well-separated. Operating under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we leveraged the fitted GLM-EIV models to approximate the Bayes boundary in the following way: we (i) sampled several hundred gene-perturbation pairs, (ii) extracted the fitted values $\hat{\beta}_g$ and $\hat{\pi}$ from the GLM-EIV models fitted to these pairs, (iii) computed the median $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ across the $\hat{\beta}_g$s and $\hat{\pi}$s, and (iv) used $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ to estimate a dataset-wide Bayes-optimal decision boundary. We repeated this procedure on both datasets, yielding a threshold of $3$ for Gasperini and $7$ for Xie. These thresholds were close to the original thresholds, which were selected in a more heuristic way.

@@

We generalized the above analysis to the entire set of positive control pairs. First, for each pair we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the magnitude of the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d). Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC). As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable. Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed CI coverage across pairs as a function of excess contamination. GLM-EIV exhibited significantly higher CI coverage than thresholded regression as the data became increasingly contaminated (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage).

@@

In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed and (optionally) impacted by the same set of confounding variables. These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens. We demonstrated through simulation studies, real data analyses, and theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings. GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells. In low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold. Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.

@@

\section{Acknowledgments and code availability}
We thank Xuran Wang helping to process the Xie dataset. We also thank Songcheng Dai for helping to deploy the GLM-EIV pipeline on Azure. This work used the Extreme Science and Engineering Discovery Environment (XSEDE; NSF grant ACI-1548562) and the Bridges-2 system (NSF grant ACI-1928147) at the Pittsburgh Supercomputing Center. Code to replicate the analyses is available at \texttt{/timothy-barry/glmeiv-manuscript}.

@@

\clearpage
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
	\caption{\textbf{Data analysis}. \textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs. \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination. Grey bands, 95\% CIs for the target of inference outputted by the methods. \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination. Colored bands, pointwise 95\% CIs for population quantities. Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
\end{figure}

@@

We performed quality control. We kept cells with percent mitochondrial reads less than 8$\%$. For the Gasperini dataset, we furthermore kept cells with UMI and gRNA counts in the 5th to 95th percentiles to reduce the effect of outliers. (We did not perform this latter filter on the Xie data, as the latter dataset appeared to be less noisy and contained fewer cells.) Next, we kept genes expressed in at least $10\%$ of cells with a mean expression level of at least $1$. We did not QC the gRNAs.

The quality-controlled Xie dataset contained $n = 101,508$ cells, $1,030$ genes, and $516$ gRNAs. There were $50,000$ \textit{in silico} negative control pairs and $681$ candidate \textit{cis} pairs. The quality-controlled Gasperini dataset contained $n = 170,645$ cells, $2,079$ genes, $6,598$ gRNAs. There were $17,028$ candidate \textit{cis} pairs, $97,818$ negative control pairs, and $322$ positive control pairs.

As a point of reference, the Gasperini dataset, which consisted of $>$115,000 pairs, took 2,120 CPU hours to analyze, or equivalently 14 hours distributed across 150 processors.

- construction of negative control pairs
- cis pairs

>>>>

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021} and improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction. We extend the classical errors-in-variables model in several directions, most importantly generalizing to exponential family-distributed response distributions and sources of measurement error.

@@

Next, we studied the thresholding method from a theoretical perspective, recovering in precise mathematical terms phenomena revealed in the empirical analysis (plus several others). We worked in a simplified Gaussian setting. Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
\begin{equation}\label{theoretical_model}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i; \quad
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
p_i \sim \textrm{Bern}(\pi); \quad
\ep_i, \tau_i \sim N(0,1),
\end{equation}
where $p_i, \tau_i,$ and $\ep_i$ are independent. For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e. $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We derive the almost sure limit of $\hat{\beta}^m_1$: 
\begin{proposition}\label{prop:convergence}
	The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
	\begin{equation}\label{thresh_est_intercepts}
	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
	\end{equation} where 
	$$
	\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
\end{proposition}
The function $\gamma: \R^4 \to \R$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \R^4 \to \R$ of $\hat{\beta}^m_1$ is given by
$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
Having derived an exact expression for  the asymptotic relative bias of  $\hat{\beta}^m_1$, we can prove several results about this quantity. For the sake of simplicity, we consider a slightly specialized setting in which we set $\pi$ to to a specific value (see Appendix \ref{sec:appendix_theory} for details; proofs and detailed statements likewise deferred to Appendix \ref{sec:appendix_theory}).

@@

To answer this question, we introduce GLM-EIV (GLM-based errors-in-variables), a method that models the entire data-generating process in a single-cell CRISPR screen experiment.

\subsection{Model}\label{sec:model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_2}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as $
p_i \sim \textrm{Bern}(\pi),$ where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}), (\ref{glmeiv_model_2}), and the marginal distribution of $p_i$ define the negative binomial GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors. See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts that may be applicable in some settings.

@@

We see from (\ref{marginal_density}) that the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities. Additionally, the GLM-EIV model is a generalization the simple errors-in-variables model (when the predictor is binary); the latter is defined as follows:
\begin{equation}\label{classical_eiv}
y_i = \beta_0 + \beta_1 x^*_i + \ep_i; \quad
x_i = x^*_i + \tau_i,
\end{equation}
where, $x^*_i \sim \textrm{Bern}(\pi), \ep_i, \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent. GLM-EIV extends (\ref{classical_eiv}) in at least three directions: first, GLM-EIV allows $y_i$ and $x_i$ to follow exponential family (i.e, not just Gaussian) distributions; second, GLM-EIV allows $y_i$ and $x_i$ to be related to $x^*_i$ through linear \textit{or} nonlinear link functions; and finally, GLM-EIV allows confounders $z_i$ to impact both $x_i$ and $y_i$. Therefore, $x_i$ and $y_i$ are dependent \textit{even after conditioning on $x^*_i$}, enabling GLM-EIV to capture more complex dependence relationships between $x_i$ and $y_i$ than is possible in (\ref{classical_eiv}) or other standard measurement error models.

@@

Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV and the thresholding method to analyze the entire Gasperini and Xie datasets. (We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well-separated. Operating under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we leveraged the fitted GLM-EIV models to approximate the Bayes boundary in the following way: we (i) sampled several hundred gene-perturbation pairs, (ii) extracted the fitted values $\hat{\beta}_g$ and $\hat{\pi}$ from the GLM-EIV models fitted to these pairs, (iii) computed the median $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ across the $\hat{\beta}_g$s and $\hat{\pi}$s, and (iv) used $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ to estimate a dataset-wide Bayes-optimal decision boundary. We repeated this procedure on both datasets, yielding a threshold of $3$ for Gasperini and $7$ for Xie. These thresholds were close to the original thresholds, which were selected in a more heuristic way.

@@

We generalized the above analysis to the entire set of positive control pairs. First, for each pair we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d). Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; upper and lower bands indicate the pointwise interquartile range of the REC). As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable. Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed CI coverage across pairs as a function of excess contamination. GLM-EIV exhibited significantly higher CI coverage than thresholded regression as the data became increasingly contaminated (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage).

@@

In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed and potentially impacted by the same set of confounding variables. These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens. We demonstrated through simulation studies, real data analyses, and theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings. GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells. On the other hand, in low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold. GLM-EIV thereby neutralizes a tuning parameter that, until this point, has been selected using heuristic procedures, with little confidence that the choice is near optimal. Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.

@@

\section{Acknowledgments and code availability}
We thank Xuran Wang helping to process the Xie dataset. We also thank Songcheng Dai for helping to deploy the GLM-EIV pipeline on Azure. This work used the Extreme Science and Engineering Discovery Environment (XSEDE; NSF grant ACI-1548562) and the Bridges-2 system (NSF grant ACI-1928147) at the Pittsburgh Supercomputing Center. Code for the analyses is available at \texttt{/timothy-barry/glmeiv-manuscript}.

@@

\clearpage
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
	\caption{\textbf{Data analysis}. \textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs. \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination. Grey bands, 95\% CIs for the target of inference outputted by the methods. \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination. Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
\end{figure}

@@

First, we performed quality control on the datasets. As is standard in single-cell analysis \cite{Choudhary2021a}, we removed cells with a high fraction ($>8\%$) of mitochondrial reads. We additionally excluded genes that were expressed in fewer than $10\%$ of cells or that had a mean expression level of less than $1$. We excluded cells in the Gasperini dataset with gene transcript UMI or gRNA counts below the 5th percentile or above the 95th percentile to reduce the effect of outliers. We did not repeat this latter quality control step on the Xie data, because the Xie data were less noisy. The quality-controlled Gasperini and Xie datasets contained $n = 170,645$ (resp. $n = 101,508$) cells, $2,079$ (resp. $1,030$) genes, and $6,598$ (resp. $516$) gRNAs.

The Gasperini dataset included $17,028$ candidate \textit{cis} pairs, $97,818$ negative control pairs, and $322$ positive control pairs. The \textit{cis} pairs consisted of genes paired to nearby enhancers with unknown regulatory effects. The negative control pairs consisted of nontargeting gRNAs paired to genes. The Xie data contained $681$ candidate \textit{cis} pairs and, unfortunately, no positive control or negative control pairs. We therefore constructed a set of $50,000$ \textit{in silico} negative control pairs on the Xie data by pairing perturbations to genes on different chromosomes; see \textit{Methods} section of \cite{Barry2020} for details.

We modeled the gene expression counts using a negative binomial distributions with unknown size parameter $\theta$; we estimated $\theta$ using the {glm.nb} package. Choudhary and Satija \cite{Choudhary2021a} report that Poisson models accurately capture highly sparse single-cell data. Although Choudhary and Satija did not investigate gRNA data specifically, we modeled the gRNA counts using Poisson distributions, as the gRNA modality exhibited greater sparsity than the gene modality.

We applied GLM-EIV and the thresholding method to analyze the entire set of pairs in both datasets. We did not report results on the candidate \textit{cis} pairs in the text because we do not know the ground truth for these pairs, making them less useful for method assessment. We focused our attention instead on the negative control pairs in both datasets and the positive control pairs in the Gasperini dataset (Figures \ref{thresholding_empirical} and \ref{fig:real_data}).

Finally, we describe in more detail how we conducted the ``excess background contamination'' analysis (Figure \ref{fig:real_data}, panels c-f). For each positive control pair, we varied excess background contamination over the grid $[0.0, 0.05, 0.1, \dots, 0.4].$ For a given level of excess background contamination, we generated $B = 50$ synthetic gRNA datasets, holding fixed the raw gene expressions, covariates, library sizes, and fitted perturbation probabilities. We fitted GLM-EIV and the thresholding method to the data, yielding estimates $[\hat{\beta}^m_1]^{(1)}, \dots, [\hat{\beta}^m_1]^{(B)}$. Next, we averaged over the $[\hat{\beta}^m_1]^{(i)}$s to obtain the mean estimate for a given pair and level of background contamination, and we calculated the REC using these mean estimates. Finally, we computed CI coverage in a somewhat nonstandard way: for a given pair and level of background contamination, we averaged over the upper and lower CIs of the $[\hat{\beta}^m_1]^{(i)}$s and checked for overlap with the raw estimate. In other words, we did not check whether the CI of each \textit{individual} $[\hat{\beta}^m_1]^{(i)}$ overlapped the ground truth; this is because we did not generate the data i.i.d., and so we cannot expect standard coverage guarantees to obtain. CI coverage (in the sense described here) for excess background contamination equal to $0$ by definitionis $1$, as depicted in Figure \ref{fig:real_data}f.
